On the spatial aggregation of ecosystem condition indicators

### 2024-11-20 manuscript/manuscript.qmd
fix typos

Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
Variables can be spatially aggregated, for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
Variables can also be normalised and turned into normative indicators for ecosystem condition.
The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
The order of these steps can also change.
Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
Variables can be truncated to produce a bound indicator scale between 0 and 1.
Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
When a variable is normalised in this way, its interpretation changes from descriptive to normative.
The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference leveles.
The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).

@@

This example is from Norway, where wolves are extinct in most of the country.
Therefore, most carnivore management areas will get an indicator value of 0.
Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
In the Norwegian Natur Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.

\*Conclusion\*: Indicators should make intuitive sense to the readers.
However, there are sometimes good reasons for making things complicated.
In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common missunderstanding about what indicator values represenets can be exaplin just one time, and not for each indicator spesifically.
It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
Wolf numbers could in this example be converted to wolf density.

@@

*Variable*: Horizontal coverage (vertical projection) of billberry (Vaccinum myrtillus) recorded in permanent vegetation plots.

@@

*Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
The data comes from nature type monitoring with reference to individual occurences (polygons).

@@

In indicator documentation or ECA repports, authors should make an effort to make a precise description of the steps in the aggregation process.
To do this we suggest using the terms as described in @fig-term.
The description can be made even easier by referencing specific aggregation pathways in @fig-pathways by number.
Besides making the indicator workflow itself more transparrent and reproduciable, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
It will make the interpretation of assessment less prone to misunderstanding.
It will also make troubleshooting easier, as i the billberry example where researchers had to spend several days looking though old code to identify the reason why two assessment produces dissimilar indicator values from the same underlying data.

@@

It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, but internal consistency within assessments should be possible.
As we saw in the wolves example (@sec-wolves), this eases interpretation and reduces the change that the meaning of indicators are misunderstood.

@@

Indicators (and variables) often exist in multiple version, varying slightly in the raw data or in the methods used to produce the data or metrics.
Yet different versions are often refered to by the same common name.
This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
We recommend making use of stable indicator IDs, unique for each version of the indicators sharing the same common name.
This has for example been implemented in [ecRxiv](https://github.com/NINAnor/ecRxiv), a publishing platform for ecosystem condition indicators.

>>>>

Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
Variables can be spatially aggregated, for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
Variables can also be normalised and turned into normative indicators for ecosystem condition.
The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
The order of these steps can also change.
Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
Variables can be truncated to produce a bound indicator scale between 0 and 1.
Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
When a variable is normalised in this way, its interpretation changes from descriptive to normative.
The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference levels.
The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).

@@

This example is from Norway, where wolves are extinct in most of the country.
Therefore, most carnivore management areas will get an indicator value of 0.
Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.

\*Conclusion\*: Indicators should make intuitive sense to the readers.
However, there are sometimes good reasons for making things complicated.
In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explain just one time, and not for each indicator specifically.
It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
Wolf numbers could in this example be converted to wolf density.

@@

*Variable*: Horizontal coverage (vertical projection) of billberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.

@@

*Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
The data comes from nature type monitoring with reference to individual occurrences (polygons).

@@

In indicator documentation or ECA reports, authors should make an effort to make a precise description of the steps in the aggregation process.
To do this we suggest using the terms as described in @fig-term.
The description can be made even easier by referencing specific aggregation pathways in @fig-pathways by number.
Besides making the indicator workflow itself more transparent and reproducible, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
It will make the interpretation of assessment less prone to misunderstanding.
It will also make troubleshooting easier, as i the billberry example where researchers had to spend several days looking though old code to identify the reason why two assessment produces dissimilar indicator values from the same underlying data.

@@

It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, but internal consistency within assessments should be possible.
As we saw in the wolves example (@sec-wolves), this eases interpretation and reduces the chance that the meaning of indicators are misunderstood.

@@

Indicators (and variables) often exist in multiple version, varying slightly in the raw data or in the methods used to produce the data or metrics.
Yet different versions are often refereed to by the same common name.
This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
We recommend making use of stable indicator IDs, unique for each version of the indicators sharing the same common name.
This has for example been implemented in [ecRxiv](https://github.com/NINAnor/ecRxiv), a publishing platform for ecosystem condition indicators.
